---
title: "Final Code"
output:
  html_document: default
  pdf_document: default
---

```{r}
library(tidyverse)
library(DescTools)
library(estimatr)
library(fastDummies)
library(ggplot2)
# read in the csv and choose the columns we will use
data <- read.csv('endline.csv', header = TRUE)
data <- data[, c("grade", "female", "age", "language", "region", "school_type", 
                 "encrypted_school_code", "treatment", "read_comp_score", "orf")]

#school_type is collinear with treatment. Have to drop it.Grouping all controls regardless of type under "control" makes this variable useless.
data <- data %>% select(-school_type)

# change the treatment variable to a binary integer
data$treatment <- ifelse(data$treatment == "Control", 0, 1)

# change school code to be 1,...,num_schools, remove original column
data$school_id <- as.numeric(factor(data$encrypted_school_code, levels = unique(data$encrypted_school_code)))
data <- data %>% select(-encrypted_school_code)

# convert character columns to factor
data[sapply(data, is.character)] <- lapply(data[sapply(data, is.character)], as.factor)

# remove rows with NAs
data <- na.omit(data)

# Age has a skewed distro with a massive outlier (someone aged 100). Will drop outlier as it seems like an error
data <- data[data$age <= 60, ]

# set up covariates (X), response (Y), and treatment (Z)
X <- data[, c("grade", "female", "age", "language")]
Z <- data$treatment
Y <- data$read_comp_score # can also use orf (oral reading fluency)
```

# Baseline school-level analysis
```{r baseline}
school_level_data <- data %>% 
  group_by(school_id) %>% 
  summarise(
    y = mean(read_comp_score, na.rm = TRUE),
    z = mean(treatment, na.rm = TRUE)        
  )

y <- school_level_data$y
z <- school_level_data$z

model <- lm(y ~ z, data = school_level_data)
model_summary_r_comp <- summary(model)

print("--- Reading Comp Results ---")
model_summary_r_comp$coefficients["z", ]
```

# School Level Data Distribution

```{r school level dist}
ggplot(school_level_data, aes(x = y, fill = as.factor(z))) +
  geom_density(alpha = 0.5) + # alpha adds transparency
  labs(title = "Average Reading Comprehension Density of Treatment and Control Schools",
       x = "Average Reading Comprehension Score",
       y = "Density",
       fill = "Treatment") + # Customize labels
  theme_classic() # Use a clean theme
```

# School Level Covariate Balance

```{r school covariate balance}
school_summary <- data %>%
  group_by(school_id) %>%
  mutate(z = Mode(treatment)) %>%
  mutate(num_2grade = sum(grade == 2)) %>%
  mutate(num_3grade = sum(grade == 3)) %>%
  mutate(num_female = sum(female == "Female")) %>%
  mutate(num_male = sum(female == "Male")) %>%
  mutate(avg_age = mean(age)) %>%
  mutate(sd_age = sd(age)) %>%
  mutate(common_language = Mode(language)) %>%
  select(-grade, -female, -age, -language, -treatment, -read_comp_score, -orf, -region) %>%
  distinct()

school_summary <- dummy_cols(school_summary, 
                             select_columns = "common_language", 
                             remove_first_dummy = TRUE) %>% select(-common_language)

covariates <- setdiff(names(school_summary), c("school_id", "z"))

balance_table <- data.frame(
  covariate = covariates,
  smd = sapply(covariates, function(v) {
    x1 <- school_summary[school_summary$z == 1, v, drop = TRUE]
    x0 <- school_summary[school_summary$z == 0, v, drop = TRUE]
    (mean(x1, na.rm = TRUE) - mean(x0, na.rm = TRUE)) /
      sqrt((var(x1, na.rm = TRUE) + var(x0, na.rm = TRUE))/2)
  })
)

ggplot(balance_table, aes(x = smd, y = covariate)) +
  geom_point(size = 3) +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dashed") +
  xlim(c(-0.2, 0.2)) + 
  labs(
    title = "Covariate Balance Between Treatment and Control",
    x = "Standardized Mean Difference",
    y = "Covariate"
  ) +
  theme_classic()
```


# Student-level analysis with Lin's estimator
```{r lin_estimator_analysis}
##OHE language
X_enc <- dummy_cols(X, 
                   select_columns = "language", 
                   remove_first_dummy = TRUE)
X_enc$female <- ifelse(X_enc$female =="Female",1,0)
#keeping only covars
X_enc <- X_enc %>% 
  select(-language) 

# Center the covariates
X_enc_cen <- scale(X_enc, center = TRUE, scale = FALSE)

X_centered <- as.data.frame(X_enc_cen)

analysis_df <- data.frame(
  read_score = Y, 
  z = Z,                             
  school_id = data$school_id,        
  X_centered                         
)

#Create regression formula dynamically
covariate_names <- colnames(X_centered)
rhs_formula <- paste("z * (", paste(covariate_names, collapse = " + "), ")")

#Model for Reading Score
formula_read <- as.formula(paste("read_score ~", rhs_formula))
lin_model_read <- lm_robust(formula_read, data = analysis_df,clusters =school_id)

# results
print("--- Lin's Estimator Results (Reading Score) student-level ---")
summary(lin_model_read)
```

# Student Level Distribution

```{r student distribution}
ggplot(analysis_df, aes(x = read_score, fill = as.factor(z))) +
  geom_bar(aes(y = after_stat(prop), group = as.factor(z)),
           position = "dodge") +
  labs(title = "Reading Comprehension Scores for Treatment and Control Students",
       x = "Reading Comprehension Score",
       y = "Proportion of Students",
       fill = "Treatment") + 
  scale_x_continuous(breaks = 0:5) + 
  theme_classic()
```


#Causal forests analysis

*Causal forests handles cross fitting and hyper parameter tuning so we only need to train the model. It also builds the propensity score and outcome models simultaneously 
```{r causal_forest_training}
library(grf)
school <- data$school_id
cf2 <- causal_forest(
  X = X_centered, 
  Y = Y, 
  W = Z,
  clusters = school, 
  tune.parameters = "all"
)
```

```{r causal_forest_analysis}
print("----Results for Reading Comp----")
ate_result_read_comp <- average_treatment_effect(cf2, target.sample = "all")
print(ate_result_read_comp)
blp_results <- best_linear_projection(cf2, X_centered)

# 2. Extract Estimates (Column 1) and Standard Errors (Column 2)
estimates <- blp_results[, 1]
std_errors <- blp_results[, 2]

# 3. Calculate Lower and Upper Bounds (Using 1.96 for 95% CI)
lower_ci <- estimates - 1.96 * std_errors
upper_ci <- estimates + 1.96 * std_errors

# 4. Combine into a clean table
results_with_ci <- cbind(blp_results, Lower_CI = lower_ci, Upper_CI = upper_ci)

# View the final table
print(results_with_ci)
```
```{r}
# Compute the ATE (uses Doubly Robust / AIPW estimation by default)
ate_result <- average_treatment_effect(cf2)

# Extract estimate and standard error
estimate <- ate_result["estimate"]
std_error <- ate_result["std.err"]

# Calculate t-value, p-value, and 95% Confidence Intervals
t_value <- estimate / std_error
p_value <- 2 * (1 - pnorm(abs(t_value))) # Two-sided p-value
ci_lower <- estimate - 1.96 * std_error
ci_upper <- estimate + 1.96 * std_error

# Combine into a clean table
ate_summary <- data.frame(
  Estimate = estimate,
  Std_Error = std_error,
  t_value = t_value,
  p_value = p_value,
  CI_Lower_95 = ci_lower,
  CI_Upper_95 = ci_upper
)

# Print the ATE Table
print("--- Average Treatment Effect (ATE) Summary ---")
print(ate_summary)
print("--- Calibration Test Results ---")
# This tests if the ATE is correct (mean.forest.prediction)
# and if the heterogeneity is real (differential.forest.prediction)
test_calibration(cf2)
```
*Results are similar to those found utilizing lin's estimator at the student level. 

